Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-10-17 article/ms.tex
add cost figure
werds

The contributions of this paper are:
\begin{itemize}
\item Solving the galactic archaeology (GA) problem using mixture modelling
% bear in mind or don't forget that GA will mean `genetic algorithms' to many readers
\item Minimum Message Length (MML) formulation
\item 2 search methods
\end{itemize}
\section{Galactic Archaeology}
% "Galaxy produces ‘molecular forests’"
% https://CosmosMagazine.com/space/galaxy-produces-molecular-forests
Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
of our galaxy by identifying ancient star formation events. Most stars are 
born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
share the same chemical composition of the gas cloud that they form from.
Stars in clusters often do not remain gravitationally bound, and disperse 
throughout the disk of our galaxy within about $10^9\,{\rm yr}$. %citation needed.
As a consequence, where a star is now does not reflect its formation site, 
which complicates our understanding of galaxy formation and evolution. However, the
observable chemical composition of a star remains largely the same throughout
that star's lifetime, providing an immutable marker (a chemical tag, or
`fingerprint') that carries with it information about where and when every star
formed.

@@

\subsection{The Message-Breaking Method} \label{sec:mbm}
The simplest strategy we introduce here, \emph{The Message-Breaking Method}, does not use any predictions of the message length of future mixtures. Instead it starts from 
a $K=1$ mixture and iteratively splits the component with the largest message length until 
the total message length no longer improves (Algorithm~\ref{alg:MBM}). This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split components (instead of splitting, merging, and deleting all components). Similarly, in \emph{The Message-Breaking Method} we only split the component with the largest message length instead of exhaustively trying to split all components at every mixture.

@@

\andy{Should we move/remove the following paragraph?}
One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. However, as we show in Section \ref{sec:results}, this hypothetical situation was not observed in any of the $\todo{5000}$ generated datasets.

\subsection{Component-Jumping Method}

The \emph{Component-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.

@@

To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets. As the output of the algorithms depends on the starting seed, and some of the methods can be affected by randomness, we perform 50 runs for each search scheme and dataset.

\andy{The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms. The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.}

@@

\todo{How we generated data sets}

\todo{Range of generated data sets}

\todo{Figure: predictions for future mixtures}

\todo{Figure: computational cost of various methods}

@@

Combining the expressions for the message length (or information content) of
each component, we arrive at our objective function $I$ to minimise across 
all model parameters (including the number of mixtures, $K$):

\begin{equation}
\todo{I = 0}
\end{equation}

@@

In Eq. \ref{eq:I1} it can be seen that for any finite mixture, $I_1$ can be
almost directly computed just from knowing the number of mixture components
$K$ and the properties of the data: the number of data points $N$ and its
dimensionality $D$. The only ambiguous term is $\sum_{k=1}^{K}\log\weight_k$,
which will depend on the data. However, for any finite mixture of $K$
components we can consider the two extremes of $\sum_{k=1}^{K}\log\weight_k$:

\begin{enumerate}
    \item The weights are distributed uniformly such that $w_k = 1/K$ and
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = \sum_{k=1}^{K}\log{\frac{1}{K}} =-K\log{K} \quad .
          \end{equation}
    \item The weights are distributed such that $K - 1$ components each 
          describe just two data points $w_{1..K-1} = 2/N$, and the remaining 
          component describes all other data points $w_{K} = 1 - 2(K-1)/N$
          as $\sum_{k=1}^{K}\weight_k = 1$. At this extreme,
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = (K-1)\log\frac{1}{N} + \log\left(1 - \frac{K - 1}{N}\right)
          %\end{equation}
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2(K-1)\log\frac{2}{N} 
                		    + \log\left(1 - \frac{2(K - 1)}{N}\right)
          \end{equation}

          \noindent{}and by making use of the identity 
            $\log{(a-b)} = \log{a} + \log{(1 - \frac{b}{a})}$:
          \begin{equation}
                \sum_{k=1}^{K}\log\weight_k = 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \quad .
          \end{equation}
          %\begin{equation}
          %      \sum_{k=1}^{K}\log\weight_k = \log{(N-K+1)} - K\log{N} \quad .
          %\end{equation}
\end{enumerate}

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
\begin{eqnarray}
\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
\end{eqnarray}
\begin{equation}
-K\log{K} \quad \leq \quad \sum_{k=1}^{K}\log\weight_k \quad \leq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}

\noindent{}which gives us analytic expressions for the upper and lower value 
of $I_1$, regardless of the values of $\vec{\weight}$, for any mixture, and
bounds on $I_{1}$ can be derived from just $\{K, N, D\}$.

>>>>

The contributions of this paper are:
\begin{itemize}
\item Solving the galactic archaeology (GA) problem using mixture modelling
% bear in mind or don't forget that GA will mean `genetic algorithms' to many readers
\item Minimum Message Length (MML) formulation
\item $\mathcal{O}\left(K\right)$ search methods for gaussian mixture  modelling
\end{itemize}
\section{Galactic Archaeology}
Galactic Archaeology~\cite{freeman2002new} aims at reconstructing the history 
of our galaxy by identifying ancient star formation events. Most stars are 
born in clusters (of order $10^0$ to $10^6$ stars at a time), and all approximately
share the same chemical composition of the gas cloud that they form from.
Stars in clusters often do not remain gravitationally bound, and disperse 
throughout the disk of our galaxy within about $10^9\,{\rm yr}$. %citation needed.
As a consequence, where a star is now does not reflect its formation site, 
which complicates our understanding of galaxy formation and evolution. However, the
observable chemical composition of a star remains largely the same throughout
that star's lifetime, providing an immutable marker (a chemical tag, or
`fingerprint') that carries with it information about where and when every star
formed.

@@

\subsection{The Message-Breaking Method} \label{sec:mbm}
The simplest strategy we introduce here, \emph{The Message-Breaking Method}, 
is a direct improvement upon the Benchmark Method that does not use predictions of the message length of future mixtures. Instead the algorithm begins with 
a $K=1$ mixture and iteratively splits the component with the largest message length until 
the total message length no longer improves (Algorithm~\ref{alg:MBM}). This approach is related to the benchmark method in that we iteratively move from a $K=1$ mixture, but we only split the component with the largest message length instead of splitting, merging, and deleting all components (and running EM at each operation).

@@

One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. In the code implementation that we release with this work, we have implemented a \emph{Strict Message-Breaking Method} variant that allows the user to specify the number of successive $K$ trials that increased the message length, before adopting the mixture with the shortest message length. By default the Strict Message-Breaking Method requires five successive mixtures where the message length did not improve.

\subsection{Message-Jumping Method}

The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of model evaluations.

@@

To evaluate the proposed search algorithms, we conduct a set of experiments on synthetic and real datasets.

@@

We set the data dimensionality $D = 2$ after previous experiments found that the data dimensionality had only a small effect on the differential cost on the search strategies tested. We chose $N$ to vary uniformly in log space from $N = 10$ data points to $N = 10^{4.5}$. For each $N$ we set $K = N/10$, and then used these values of $N$, $D$, and $K$ to generate a synthetic dataset. The weights for each mixture were drawn from a Dirichlet distribution with concentration $\alpha = 100$ while ensuring that $\weight_k \in \left(\frac{2}{N}, 1-\frac{2(K-1)}{N}\right)$. The means for each component were drawn from $\mathcal{U}\left(-10, +10\right)$. The correlation coefficients  for the component covariance matrices were drawn from
\begin{equation}
	\rho \sim \frac{\left[1-\mathcal{U}(-1, +1)^2\right]^{\frac{1}{2}(s - 4)}}{\mathcal{B}\left(\frac{1}{2}, \frac{1}{2}(s - 4)\right)}
\end{equation}
\noindent{}where $s$ is an isotropy hyperparameter that describes the probability of correlations between data dimensions, which we fix as $s=100$. The diagonal variances for the component covariance matrices were drawn from $|\mathcal{N}\left(0, 0.1\right)|$. Given these randomly drawn parameters we generated $N\weight_k$ members for each component, rounding as necessary to ensure the total number of generated data points is kept at $N$.

We evaluated the following methods on all generated datasets: the benchmark method (Algorithm \ref{alg:kasarapu}), the Message-Breaking Method (Algorithm \ref{alg:MBM}), and \todo{X}. The cost (in seconds) is shown as a function of the number of data points $N$ and true number of clusters $K$ in Figure \ref{fig:algorithm_costs}.
\todo{Cost is approximately linear for MBM, but of O(3) for kasarpu. Not surprising because Kasarapu is a greedy search method, requiring at least 3*K operations (split, merge, delete) at each mixture.}

We considered the search algorithm to be converged to the global minima if the calculated message length is shorter than the message length given the true mixture parameters. \todo{Something about when converged and not}

\todo{Figure: predictions for future mixtures for one generated data set}

\andy{As the output of the algorithms depends on the starting seed, and some of the methods can be affected by randomness, we perform 50 runs for each search scheme and dataset.  The different search schemes are validated using the Kolmogorov-Smirnov (KS) non-parametric test~\citep{Pettitt77S}, which checks for a statistical significance in the difference between the performances of the algorithms. The 50 runs are submitted to the KS analysis with a null hypothesis of no difference between the performances of the different search methods.}

\todo{describe APOGEE data}

@@

Combining the expressions for the message length (or information content) of
each component, we arrive at our objective function $I$ to minimise across 
all model parameters (including the number of mixtures, $K$):
\begin{eqnarray}
	I(\vec{\data}|\vec{\psi}, N, D) &=& -\log\mathcal{L}\left(\vec{\data}|\vectheta\right) - \frac{(D + 2)}{2}\sum_{k=1}^{K}\log{|{\cov}_k|} \nonumber \\
	&& \cdots\ + \left(\frac{D(D+3)}{4} - \frac{1}{2}\right)\sum_{k=1}^{K}\log{\weight_k} + K\left(1-\frac{D}{2}\right)\log{2} \nonumber \\ && \cdots\ + \log\Gamma{(K)} + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N} \,\,\,\, . 
	\label{eq:objective_function}
\end{eqnarray}

@@
